---
title: "Data visualisation:"
subtitle: "A bridge between politics, society, and statistics"
author: "Zoe Meers"
institute: "United States Studies Centre, University of Sydney"
date: "4 May 2020"
output:
  xaringan::moon_reader:
    css: ["default", "ninjutsu",  "metropolis-fonts"]
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(ussc)
library(flipbookr)

knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina = 2, out.width="90%")
knitr::opts_hooks$set(fig.callout = function(options) {
  if (options$fig.callout) {
    options$echo <- FALSE
    options$out.height <- "70%"
    options$fig.width <- 8
    options$fig.height <- 6.75
  }
  options
})
```

class: inverse, left, middle

* Election forecasting

* Data journalism

* Political data visualisation in **R**

* Data analytics in think tanks

---
# Election forecasting: FiveThirtyEight

<iframe width="1280" height="600" src="https://projects.fivethirtyeight.com/2020-primary-forecast/" frameborder="0" allowfullscreen></iframe>

---
## Election forecasting: FiveThirtyEight

* Rolling poll averages over time

* Probabilistic forecasts during an election campaign

* Compelling, interactive displays of polls which are data points for the predictive model

---
class: inverse, center, middle

We've come a long way from this:

**"There are three kinds of lies: lies, damned lies, and statistics."** 
-[Author unknown, via Mark Twain in *Chapters from My Autobiography (1907)*]


---
## Data journalism

* Not unique to FiveThirtyEight

* Data journalism teams exist in most major newsrooms globally, including the New York Times, the Washington Post, the Financial Times, the Guardian, and so on

* In Australia the most prolific newsroom data viz teams are at the ABC and Guardian Australia

* Interactive JavaScript-based web pages - scrolly telling, **d3.js** - are key 

* **R** and/or **Python** are used for the data analysis stage, rarely for the final product (except for the BBC!)


---
class: inverse, center, middle

But you *can* create beautiful production-ready graphs in **R** as well!

---
## Visualising political data in **R**

* Some useful packages for politics & data are:
  * **fivethirtyeight**, **politicaldata**, **ggparliament**

* **ggparliament** is an R package for visualising the physical shape of legislatures

* It is a **ggplot2** extension


---
## Visualising legislatures with **ggparliament**

```{r, include = FALSE}
election_data <- ggparliament::election_data %>% 
  add_row(year = 2018, country = "USA", house = "Representatives", party_long = "Republican", party_short = "GOP", seats = 199, government = 0, colour = "#E81B23", .after = 6) %>% 
  add_row(year = 2018, country = "USA", house = "Representatives", party_long = "Democratic", party_short = "Dem", seats = 235, government = 1, colour = "#E81B23", .after = 7)
```

```{r ggparliament, include = FALSE, fig.callout=TRUE}
election_data %>%
  filter(country == "USA" &
    house == "Representatives")  %>% 
  group_split(year) %>% 
  map_dfr(
    ~ggparliament::parliament_data(.,
      party_seats = .$seats,
      parl_rows = 10,
      type = "semicircle")
    ) %>% 
  select(-c(country,
            house,
            party_long,
            seats)
         ) %>% 
  ggplot(aes(x, y)) +
  facet_wrap(~year, nrow = 2) +
  ggparliament::theme_ggparliament() + 
  ggparliament::geom_parliament_seats(
    size = 1,
    aes(colour = party_short)
    ) + 
  theme(legend.position = 'bottom') +
  ggparliament::geom_highlight_government(
    government == 1,
    size = 1.1,
    colour = "grey"
    ) + 
  labs(colour = NULL, 
       title = "American Congress",
       subtitle = "The party that has control of Congress is highlighted in grey."
       ) +
  scale_colour_manual(values = c("#E81B23","#3333FF"), 
                      limits = c("GOP", "Dem"))
```

---
## Visualising legislatures with **ggparliament**

`r chunk_reveal("ggparliament", break_type = "auto")`

---
## Visualising legislatures with **ggparliament**

```{r ggparliament_au, include = FALSE, fig.callout=TRUE}
australia <- election_data %>%
  filter(country == "Australia" &
    house == "Representatives" &
    year == 2016) 

ggparliament::parliament_data(
  election_data = australia,
    party_seats = australia$seats,
    parl_rows = 4,
    type = "horseshoe") %>% 
  ggplot(aes(x, y)) +
  ggparliament::theme_ggparliament() + 
  ggparliament::geom_parliament_seats(
    aes(colour = party_short)
    ) + 
  scale_colour_manual(values = australia$colour, 
                      limits = australia$party_short) +
  theme(legend.position = 'none') +
  ggparliament::draw_partylabels(
                   type = "horseshoe", 
                   party_names = party_short,
                   party_seats = seats, 
                   party_colours = colour) +
  ggparliament::draw_totalseats(
                   n = 150,
                   type = "horseshoe") +
  labs(colour = NULL, 
       title = "Australian House of Representatives, 2016 election"
       ) +
  coord_cartesian(clip = "off")
```
---
## Visualising legislatures with **ggparliament**

`r chunk_reveal("ggparliament_au", break_type = "auto")`

---
## Visualising legislatures with **ggparliament**


```{r, include = FALSE}
uk <- election_data %>%
  filter(country == "UK") %>% 
  mutate(party_short = case_when(
    party_short == "Con" ~ "Tories",
    TRUE ~ party_short
  ),
    party_short = fct_other(party_short,
                                 keep = c(
                                   "Lab",
                                   "SNP",
                                   "LibDem",
                                   "Tories",
                                   "DUP"
                                 )),
  colour = case_when(
           party_short == "Other" ~ "grey",
           TRUE ~ colour)
           ) %>% 
  select(-party_long) %>% 
  group_by(year, party_short, colour, government) %>% 
  tally(seats, name = "seats") %>% 
  ungroup() 

uk_2019 <- tribble(
  ~year, ~party_short, ~colour, ~seats, ~government, 
  2019, "Lab", "#DC241F", 203, 0,
  2019, "SNP", "#FEF987", 48, 0,
  2019, "Tories", "#0087DC", 365, 1,
  2019, "LibDem", "#FAA61A", 11, 0,
  2019, "DUP", "#D46A4C", 8, 0,
  2019, "Other", "grey", 15, 0
)

uk <- bind_rows(uk, uk_2019)
```


```{r ggparliament_uk, include = FALSE, fig.callout=TRUE}
uk %>%
  group_split(year) %>%
  map_dfr(~ggparliament::parliament_data(
    election_data = .,
    party_seats = .$seats,
    group = .$government,
    type = "opposing_benches")) %>%
  ggplot(aes(x, y)) +
  ggparliament::geom_parliament_seats(
    size = 1,
    aes(colour = party_short)) +  
  facet_wrap(~year, ncol = 2) + 
  scale_color_manual(values = uk$colour, 
                     limits = uk$party_short) +
  ggparliament::theme_ggparliament(
    background_colour = TRUE) +
  labs(colour = "Parties", 
       title = "UK House of Commons",
       subtitle = "The goverment sits to the Speaker's right"
       ) +
  coord_flip()
```

---
## Visualising legislatures with **ggparliament**

`r chunk_reveal("ggparliament_uk", break_type = "auto")`


---
## Data analytics in think tanks

* Data-driven policy wonks

* Urban Institue, Pew, Brookings, CSIS

* USSC, Grattan, Lowy

* Communicating for a general audience

---
## Data analytics in think tanks

* Key things to keep in mind in the think tank world:
  
  * Simple, facts-based viz

  * Make any data analysis easy to peruse

  * Consider guiding the reader and pointing out what they should be looking at 

---
## Data analytics in think tanks

* Create a common workplace-wide **ggplot2** theme, wrap it up into a package

* Have extensive documentation and guides for data visualisation and analysis pipelines

* Remember your audience

---
## Data analytics in think tanks

* Mobile-friendly graphs in **R**

* SVG/PNG/PDF

* When knitting an RMD file to a HTML page and you want graphs that are responsive to a changing browser size, set `out.width` and `out.height` instead of `fig.width` and `fig.height`

* This is particularly important when dealing with HTML widgets that do not scale well

* A quick hack is to copy and paste this into your chunk options: `dev = 'svg', out.width = "100%", out.extra='style="position:absolute; padding:10px; display: inline-block;"'`

* Other HTML arguments can be passed to `out-extra`.

---
## Data analytics in think tanks

* Some R Markdown tips for "interactive" content 

  * The use of `{.tabset}`, ` {.tabset .tabset-pills}`, and `{.tabset .tabset-dropdown}` in headers
  
  * Switching between graphs
  
  * Gives the impression of redrawing axes -- easy in **JavaScript**, hard in **R** (with the exception of Shiny)
  
---
<iframe width="1280" height="600" src="https://www.ussc.edu.au/analysis/australia-and-the-united-states-by-the-numbers" frameborder="0" allowfullscreen></iframe>


---
## Data analytics in think tanks

* Mathematical knowledge is comparatively less useful

* What matters is being able to draw out insights that are relevant for public policy

* Some combination of tools (**R**, **Python**), statistical knowledge, and substantive domain knowledge

---
## Data analytics in think tanks

* The ability to code and do data analytics is very useful for think tanks

* Rigorous analysis, consideration of missing information, uncertainty

* Quantify what works and what does not

---
## Communicating with data

* New skill set for traditionally qualitative domains

* Requires more education around statistics in the social sciences

* Particularly necessary in journalism 

---
## Communicating with data

* Why does basic numeracy and data literacy matter?

* Quantifying social phenomena is very important when it comes to things like public opinion, budgets, election campaigns, support for policy initiatives, and other events such as pandemics...

* Avoid the trap of people misusing statistics for their gain - i.e. to prove their point even if the evidence says otherwise. 

* Data visualisation helps -- takes complicated information and simplifies presentation.





